Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations

### 2021-12-07 analysis/paper/paper.Rmd
cleaned a bit of methods and code

The first objective was to estimate the population size of piscine predators in the Deadwater Slough to begin to their potential impacts on juvenile Chinook Salmon. Our initial intent was to estimate predator abundances using mark-recapture methods during fall 2019 and spring 2020 during the emigration of the 2018 brood year. We attempted to sample near the peaks of the juvenile DSR (fall) and NRR (spring) emigrations. However, due to logistical concerns during the onset of the Covid-19 pandemic, the spring 2020 sampling event was delayed until fall 2020. Although this provided the benefit of an additional abundance estimate during the DSR emigration, estimates during the NRR emigration were still lacking, and so an additional sampling event was added for spring 2021. The study design was such that the fall 2019 and fall 2020 sampling efforts included both a "mark week" and "recapture week" intended for mark-recapture models. However, due to budget constraints the spring 2021 effort only included a single "mark week"; instead, we calculated the catch-per-unit-effort (CPUE; fish per hour) for the spring 2021 sampling event for comparison to the fall efforts. In the end, sampling dates were as follows: fall 2019 (mark = November 12-14; recapture = November 19-21), fall 2020 (mark = October 20-23; recapture = October 27-30); spring 2021 (mark = May 18-21). The recapture events were held just one week following the mark events to reduce opportunity for immigration/emigration of predators between events i.e., to avoid violation of a closed population for mark-recapture models.

@@

```{r cmr-effort-data}
raw_df = readxl::read_excel(here("analysis/data/raw_data/Deadwater_predation_mark_recapture_11.09.21.xlsx")) %>%
  clean_names() %>%
  mutate(across(date,
                ymd)) %>%
  mutate(across(c(start_time, end_time),
                ~ ymd_hms(paste(year(date), month(date), day(date), hour(.), minute(.), second(.))))) %>%
  # split into sampling periods
  mutate(event_name = if_else(month(date) > 8,
                              paste("Fall", year(date), sep = "_"),
                              paste("Spring", year(date), sep = "_")))

# focus on pikeminnow
cmr_df = raw_df %>%
  filter(count > 0) %>%
  # pull out only Northern Pikeminnow
  filter(species == "Northern Pikeminnow") %>%
  mutate(mark_release = if_else(fish_status == "Live" &
                                  capture_type == "Mark",
                                T, F),
         recap = if_else(capture_type == "Recapture",
                         T, F))

effort_df = raw_df %>%
  select(event_name,
         site,
         date,
         crew,
         crew_count,
         method,
         equipment,
         start_time:total_effort,
         total_day_effort) %>%
  distinct() %>%
  mutate(across(total_effort,
                ~ . * crew_count)) %>%
  group_by(event_name,
           date) %>%
  summarise(across(c(total_effort,
                     crew_count),
                   sum),
            .groups = "drop")

cpue_df = effort_df %>%
  # add how many pikeminnow were caught
  left_join(cmr_df %>%
              group_by(event_name, species, date) %>%
              summarise(n_fish = n(),
                        n_large_fish = sum(length>=228,
                                           na.rm = T),
                        .groups = "drop")) %>%
  # calculate CPUE a few different ways
  mutate(cpue = n_fish / total_effort,
         cpue_per_day = n_fish / crew_count,
         cpue_large_per_day = n_large_fish / crew_count)

@@

```{r lincoln-petersen-estimator}
# using Lincoln-Petersen model, combining first 4 and last four days
# define mark and recapture periods
cap_periods = tibble(event_name = "Fall_2019",
                     cap_period = "Mark",
                     int = interval(start = ymd('20191112'),
                                    end = ymd('20191114'))) %>%
  bind_rows(tibble(event_name = "Fall_2019",
                   cap_period = "Recapture",
                   int = interval(start = ymd('20191119'),
                                  end = ymd('20191121')))) %>%
  bind_rows(tibble(event_name = "Fall_2020",
                   cap_period = "Mark",
                   int = interval(start = ymd('20201020'),
                                  end = ymd('20201023')))) %>%
  bind_rows(tibble(event_name = "Fall_2020",
                   cap_period = "Recapture",
                   int = interval(start = ymd('20201027'),
                                  end = ymd('20201030'))))

cmr_df$cap_period = NA_character_
for(i in 1:nrow(cap_periods)) {
  my_int = cap_periods$int[i]
  index = which(cmr_df$date %within% my_int)
  cmr_df$cap_period[index] = cap_periods$cap_period[i]
}

LP_mod = cmr_df %>%
  # filter out spring 2021 (no recapture)
  filter(event_name != "Spring_2021") %>%
  group_by(event_name, species) %>%
  summarise(M = sum(cap_period == "Mark" & fish_status == "Live", na.rm = T),
            n = sum(cap_period == "Recapture"),
            m = sum(cap_period == "Recapture" & capture_type == "Recapture", na.rm = T),
            .groups = "drop") %>%
  nest(data = -c(event_name, species)) %>%
  crossing(model = c("Petersen",
                     "Chapman")) %>%
  mutate(mr_model = map2(data,
                       model,
                      .f = function(x, y) {
                        with(x,
                             mrClosed(M = M,
                                      n = n,
                                      m = m,
                                      method = y))
                      }),
         N_summ = map(mr_model,
                      .f = summary,
                      incl.SE = T),
         CI = map(mr_model,
                  .f = confint)) %>%
  mutate(N = map_dbl(N_summ,
                     .f = function(x) x[1]),
         SE = map_dbl(N_summ,
                      .f = function(x) x[2]),
         Lci = map_dbl(CI,
                         .f = function(x) x[1]),
         Uci = map_dbl(CI,
                         .f = function(x) x[2])) %>%
  select(-N_summ, -CI)

@@

```{r spring-abund}
total_cpue = cpue_df %>%
  mutate(across(event_name,
                ~ str_replace(., "_", " "))) %>%
  group_by(species, event_name) %>%
  summarise(across(c(total_effort, 
                     n_fish,
                     n_large_fish,
                     angler_days = crew_count),
                   sum),
            .groups = "drop") %>%
  mutate(across(total_effort,
                ~ . / 60)) %>%
  mutate(cpue = n_fish / total_effort,
         cpue_per_day = n_fish / angler_days,
         cpue_large_per_day = n_large_fish / angler_days)

spring_abund = all_est %>%
  select(-range, -prop_range) %>%
  inner_join(total_cpue,
             by = c("event_name", "species")) %>%
  select(species, model, 
         model_event = event_name,
         N_F = N, 
         Lci_F = Lci,
         Uci_F = Uci,
         cpue_F = cpue) %>%
  full_join(total_cpue %>%
              filter(str_detect(event_name, "Spring")) %>%
              select(species,
                     cpue_S = cpue),
            by = "species") %>%
  mutate(N_S = N_F * (cpue_S / cpue_F),
         Lci_S = Lci_F * (cpue_S / cpue_F),
         Uci_S = Uci_F * (cpue_S / cpue_F))

@@

```{r cpue-fig, fig.cap = "Total catch per unit effort across entire sampling event."}
total_cpue %>%
  ggplot(aes(x = event_name,
             y = cpue_large_per_day,
             fill = event_name)) +
  geom_col() +
  scale_fill_brewer(palette = "Set1",
                    name = "Event") +
  labs(x = "Event",
       y = "CPUE (Large Pikeminnow caught per angler day)") +
  theme(legend.position = "none")

>>>>

Our first objective was to estimate the population size of piscine predators in the Deadwater Slough to begin to understand their potential impacts on juvenile Chinook Salmon. The intent was to estimate predator abundances using mark-recapture methods during fall 2019 and spring 2020 during emigration of the 2018 brood year. We attempted to sample near the peaks of the juvenile DSR (fall) and NRR (spring) emigrations. However, due to logistical concerns during the onset of the Covid-19 pandemic, the spring 2020 sampling event was delayed until fall 2020. Although this provided the benefit of an additional abundance estimate during the DSR emigration, estimates during the NRR emigration were still lacking, and so an additional sampling event was added for spring 2021. The study design was such that the fall 2019 and fall 2020 sampling efforts included both a "mark week" and "recapture week" intended for mark-recapture models. However, due to budget constraints the spring 2021 effort only included a single "mark week"; instead, we calculated the catch-per-unit-effort (CPUE; fish per hour) for the spring 2021 sampling event for comparison to the fall efforts. In the end, sampling dates were as follows: fall 2019 (mark = November 12-14; recapture = November 19-21), fall 2020 (mark = October 20-23; recapture = October 27-30); spring 2021 (mark = May 18-21). The recapture events were held just one week following the mark events to reduce opportunity for immigration/emigration of predators between events i.e., to avoid violation of a closed population for mark-recapture models.

@@

```{r capture-data, eval = T}
raw_df = readxl::read_excel(here("analysis/data/raw_data/Deadwater_predation_mark_recapture_11.09.21.xlsx")) %>%
  clean_names() %>%
  mutate(across(date,
                ymd)) %>%
  mutate(across(c(start_time, end_time),
                ~ ymd_hms(paste(year(date), month(date), day(date), hour(.), minute(.), second(.))))) %>%
  # break days into "sampling events"
  mutate(event_name = if_else(month(date) > 8,
                              paste("Fall", year(date), sep = "_"),
                              paste("Spring", year(date), sep = "_")))

# focus on pikeminnow, to be used for mark-recap models
cmr_df = raw_df %>%
  filter(count > 0) %>%
  # pull out only Northern Pikeminnow
  filter(species == "Northern Pikeminnow") %>%
  mutate(mark_release = if_else(fish_status == "Live" &
                                  capture_type == "Mark",
                                T, F),
         recap = if_else(capture_type == "Recapture",
                         T, F))

@@

```{r effort, eval = T}
effort_df = raw_df %>%
  select(event_name,
         site,
         date,
         crew,
         crew_count,
         method,
         equipment,
         start_time:total_effort,
         total_day_effort) %>%
  distinct() %>%
  filter(method == "Angling") %>%
  mutate(across(total_effort,
                ~ . * crew_count)) %>%
  rename(total_effort_min = total_effort) %>%
  group_by(event_name,
           date) %>%
  summarise(across(c(total_effort_min,
                     crew_count),
                   sum),
            .groups = "drop")

# What size (mm) is considered a large pikeminnow?
large_size = 228

# catch per unit effort
cpue_df = effort_df %>%
  # add back in the number of pikeminnow caught
  left_join(cmr_df %>%
              group_by(event_name, species, date) %>%
              summarise(n_fish = n(),
                        n_large_fish = sum(length >= large_size,
                                           na.rm = T),
                        .groups = "drop")) %>%
  # calculate CPUE a few different ways
  mutate(catch_per_min = n_fish / total_effort_min,
         catch_per_hr  = n_fish / (total_effort_min / 60),
         catch_per_angler_day = n_fish / crew_count,
         catch_lg_per_angler_day = n_large_fish / crew_count)

```{r lincoln-petersen-estimator, eval = T}
# using Lincoln-Petersen model, combining first 4 and last four days
# define mark and recapture periods
cap_periods = tibble(event_name = "Fall_2019",
                     cap_period = "Mark",
                     int = interval(start = ymd('20191112'),
                                    end = ymd('20191114'))) %>%
  bind_rows(tibble(event_name = "Fall_2019",
                   cap_period = "Recapture",
                   int = interval(start = ymd('20191119'),
                                  end = ymd('20191121')))) %>%
  bind_rows(tibble(event_name = "Fall_2020",
                   cap_period = "Mark",
                   int = interval(start = ymd('20201020'),
                                  end = ymd('20201023')))) %>%
  bind_rows(tibble(event_name = "Fall_2020",
                   cap_period = "Recapture",
                   int = interval(start = ymd('20201027'),
                                  end = ymd('20201030'))))

# add cap_period column, fill w/ NA
cmr_df$cap_period = NA_character_
# fill cap_period w/ Mark or Recapture
for(i in 1:nrow(cap_periods)) {
  my_int = cap_periods$int[i]
  index = which(cmr_df$date %within% my_int)
  cmr_df$cap_period[index] = cap_periods$cap_period[i]
}

# Lincoln-Peterson Model
LP_mod = cmr_df %>%
  # filter out spring 2021 (no recapture)
  filter(event_name != "Spring_2021") %>%
  group_by(event_name, species) %>%
  summarise(M = sum(cap_period == "Mark" & fish_status == "Live", na.rm = T),
            n = sum(cap_period == "Recapture"),
            m = sum(cap_period == "Recapture" & capture_type == "Recapture", na.rm = T),
            .groups = "drop") %>%
  nest(data = -c(event_name, species)) %>%
  crossing(model = c("Petersen",
                     "Chapman")) %>%
  mutate(mr_model = map2(data,
                       model,
                      .f = function(x, y) {
                        with(x,
                             FSA::mrClosed(M = M,
                                          n = n,
                                          m = m,
                                          method = y))
                        }),
         N_summ = map(mr_model,
                      .f = summary,
                      incl.SE = T),
         CI = map(mr_model,
                  .f = confint)) %>%
  mutate(N = map_dbl(N_summ,
                     .f = function(x) x[1]),
         SE = map_dbl(N_summ,
                      .f = function(x) x[2]),
         Lci = map_dbl(CI,
                         .f = function(x) x[1]),
         Uci = map_dbl(CI,
                         .f = function(x) x[2])) %>%
  select(-N_summ, -CI)

@@

```{r spring-abund}
total_cpue = cpue_df %>%
  mutate(across(event_name,
                ~ str_replace(., "_", " "))) %>%
  group_by(species,
           event_name) %>%
  summarise(across(c(total_effort_min,
                     n_fish,
                     n_large_fish,
                     angler_days = crew_count),
                   sum),
            .groups = "drop") %>%
  mutate(catch_per_min = n_fish / total_effort_min,
         catch_per_hr = n_fish / (total_effort_min / 60),
         catch_per_angler_day = n_fish / angler_days,
         catch_lg_per_angler_day = n_large_fish / angler_days)

spring_abund = all_est %>%
  select(-range, -prop_range) %>%
  inner_join(total_cpue,
             by = c("event_name", "species")) %>%
  select(species,
         model,
         model_event = event_name,
         N_F = N,
         Lci_F = Lci,
         Uci_F = Uci,
         cpue_F = catch_per_hr) %>%
  full_join(total_cpue %>%
              filter(str_detect(event_name, "Spring")) %>%
              select(species,
                     cpue_S = catch_per_hr),
            by = "species") %>%
  mutate(N_S = N_F * (cpue_S / cpue_F),
         Lci_S = Lci_F * (cpue_S / cpue_F),
         Uci_S = Uci_F * (cpue_S / cpue_F))

@@

```{r cpue-fig, fig.cap = "Total catch per unit effort across entire sampling event."}
total_cpue %>%
  ggplot(aes(x = event_name,
             y = catch_lg_per_angler_day,
             fill = event_name)) +
  geom_col() +
  scale_fill_brewer(palette = "Set1",
                    name = "Event") +
  labs(x = "Event",
       y = "CPUE (Large Pikeminnow caught per angler day)") +
  theme(legend.position = "none")
